A Survey of Scientific Workflow Management Systems

### 2014-04-05 paper.tex
removed unnecessary stuff
some new bib entries

\begin{enumerate}
%
\item \textbf{Scalability and Optimization.} SWEs are more likely to involve extensive usage of distributed resources than conventional programs. Owing to the large-scale experiments, for which scientific users turn to automation, their demands for computational power often exceeds that of a single personal computer. Such applications are often composed of multiple, distinct computational stages making them well-suited to the workflow paradigm. These properties also make them ideal candidates for a large-scale distributed computing environment. The case for suitability of SWEs on large-scale distributed infrastructures has been well-argued upon in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to ensure scalable and optimized performance is vital.
%
\item \textbf{Data Management.} The scale of data transfer is much higher in SWEs from one execution site to others as well as among storage resources while it is lower (ranging from variables to small data structures) in programs.
%
\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).} Building SWE's interfaces to distributed systems is a complex task. The distributed nature of resource utilization and asynchronous execution of workflow requires a set of interdependent parameters to be configured correctly. Further complexity is added because of complex dataflow requirements, interactions with diverse tools and linking with legacy application codes.
%
\item \textbf{Expressiveness.} Programming languages are inherently control flow oriented where data flow is a secondary aspect. In fact, the programmer must explicitly make sure that correct the inputs are available to instructions and are in the right format. A program proceeds as the execution of its statements from beginning to end and completes its execution in a predefined sequence. However, in SWEs, the flow is often determined by the data availability for execution. This model of computation is known as dataflow computing. A workflow thus does not follow a fixed order of execution and different execution sequences are possible which are dominated by dataflow. The unit of work in SWEs is at a higher level of granularity. A unit of execution is coarse-grained, and often termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE community. This notion of high-level granularity drives the depth to which a workflow can be expressed in a given language form. In the rest of this paper, this unit of work will be referred to as \textit{processor}. On the other hand, a unit of execution in most programming languages is a much simpler and finer executable statement that accomplishes a small computational task. This presents a whole new set of requirements for expressing the processors and how they should interact and operate upon data.
%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain experts to utilize the computational power. This makes it essential for them to have a friendly, and intuitive user-interface for novice and non-expert users. For this reason, SWE user interfaces are dominated by the visual paradigm of programming. On the other hand, traditional programs are most often written by expert programmers. Thus, they have a text-based composition environment albeit with visually assisting programming environments such as IDEs. However, they are seldom to the extent that a non-expert user can graphically compose programs.

@@

\section{Architecture of a Generic SWE}
Figure \ref{fig:wfms} shows a high-level architecture of a SWE. It shows key components and their interaction in a SWE. These components are the focus of interest for the present work. A \textit{user interface} interacts with the \textit{workflow engine} using commands and messages. The \textit{workflow language} becomes an intermediate representation for communication between user interface and the \textit{workflow engine}. The workflow engine acts as a compiler/interpreter and scheduler for the workflow tasks and interfaces with the underlying \textit{distributed resources}. It also monitors the workflow progress and collects the intermediate and final results.
%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing infrastructure such as clusters, supercomputers, clouds, grids. Such environments are in use today for production level work in different scientific domains worldwide. They do the basic task of linking inter-related processors of a multi-staged scientific computation. However, they differ in the features they provide based upon the scientific domain they offer service for, and the computational model they follow. Often, one of the main reasons for such difference is the specialized components and requirements for a particular domain. Owing to this fact, a given workflow system dominates a particular area of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora of SWEs suitable for general purpose workflow style computations.
% Ketanmark

@@

MOTEUR has sophisticated data management mechanisms within the workflow definition. It provides a diverse set of iterations strategies well adapted for different combinations of input datasets at the processor. An iteration strategy allows different combinations of the data arriving at a processor along its multiple inputs. Running legacy application code on grids has been eased by MOTEUR by providing a generic grid-service wrapper for these applications \cite{glatard-montagnat-etal:2008a}. Grid-services thus created can be readily deployed on the grids by the MOTEUR workflow enactor. MOTEUR workflows are written in the scufl language and are compatible with those written for the Taverna SWE. It also supports an extension of scufl language called the Gwendia language, developed at the same laboratory. 
\paragraph{Makeflow} Business Process Execution Language (BPEL) \cite{oasis:2007} is a workflow language pioneered by IBM Corporation and used predominantly to define business workflows. It is now a broad standard for business workflow applications. BPEL is an open language and commercial enactors are available that provide enactment functionality for the BPEL workflows. There has been attempts to use BPEL or its extended version in the scientific applications \cite{slominski:2007,bpel}. However, the language does not suit well to the data-intensive, grid workflows and has a limited adoption in the scientific community \cite{tan-missier-etal:2010,barga-gannon:2007,scherp-hoing-etal:2009,emmerich-butchart-etal:2005}. Apart from that, a lack of an open-source workflow engine with graphical interface also contributes to its limited acceptance in academic research. However, BPEL has been famously unsuitable for data intensive scientific applications  BPEL lacks different user-defined strategies to effect multiple invocations of each processor according to the application semantics. BPEL also lacks data streams and implicit parallelism forcing a user to explicitly define parallelism using one of its parallel processing constructs such as \texttt{foreach}.
\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based SWE developed at the Cardiff University. It consists of a workflow composition GUI, a framework to integrate workflow with the different execution environments and middlewares and a set of predefined services written as java components. A Triana workflow is expressed as an XML grammar similar to a WSDL specification for webservices \cite{taylor-wand-etal:2005}. A highlight of Triana is that it provides ready-to-use interface for a wide variety of middleware and distributed infrastructures including P2P, grids and web-services frameworks. A Triana workflow provides dataflow as well as control flow support. Triana allows conditional branching and looping in its workflows.
\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at the University of California. Kepler workflows follow an actor-director metaphor where an actor represents a unit of execution and the director represents the paradigm for executing the workflows consisting of the actors. Depending upon the types of directors, a Kepler workflow may exhibit a different type of execution behavior. A workflow designer can chose from among directors based upon the purpose of a workflow. Few of these include Process Network (PN) for a parallel execution at the workflow level, Synchronous Data Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its workflow representation. MoML is an extended XML syntax providing a rich vocabulary to compose Kepler workflows. A workflow processor is called an `actor' in Kepler. 
\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the University of North Carolina, Wilmington, facilitates a graphical composition and execution of DAG-based workflows. Like Kepler, GridNexus is also based on Ptolemy II framework. GridNexus workflows are expressed in an XML-based scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a customizable and extensible scripting language allowing users to write their own workflow processor types. GridNexus workflows are grid oriented and can be deployed on grid resources accessed through a persistent WS-RF (Web Services Resource Framework) interface.
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed computing framework that can map abstract workflows as DAGs to the concrete underlying resources using intelligent and dynamic scheduling policies. A Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in XML). DAX describes the logical workflow components, their inputs and outputs and their links and dependencies in a Pegasus workflow. These kind of workflows has been shown to enact optimally using Pegasus' smart memory and disk management techniques. Pegasus manager buffers and maintains a fix-sized window of executable tasks before accepting new tasks.

@@

% Put this somewhere else, either in the analysis or may be in the conclusions and summary.
%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language developed as a part of the current work is similar in spirit to the SwiftScript. Similar to SwiftScript, gscript adopts futures to express asynchronous execution in a script-based language. A SwiftScript workflow contains statically typed datasets that can be specified using ``annotations'' in the script. Unlike SwiftScript, gscript does provide user-defined strategies in order to effect data combinations in the workflow. SwiftScript provides for fault-tolerance and exception handling at run-time which is lacking currently in gscript. The underlying datasets organization in swift is represented using an XDTM XML-representation. This means that effectively, a workflow is SwiftScript plus the XDTM data representation. A workflow is simpler and compact in gscript as its data description is embedded within a workflow representation. SwiftScript supports creation of nested-workflows in the form of compound procedures. Conditional and Loop execution is supported using the if and for/while statements in both the languages.
\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE) \cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck. ASKALON is described as a Scientific PSE capable of composing, hosting, executing, sharing and reusing scientific workflows written in the Abstract Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is capable of expressing scientific application components chains at a high-level of abstraction. A concrete workflow consists of the abstract workflow combined with the resources and target execution environment binding the workflow. The graphical interface of AGWL is based upon the UML (Unified Modeling Language) diagrams.
\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service (GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First institute. It implements a dynamic workflow composition and enactment model based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}. GWES enables key workflow development and deployment lifecycle phases including visual composition, data-staging, Grid middleware interfacing and execution monitoring. Currently, GWES is intensively being used in the field of Medical Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany. It has been successfully interfaced with the German D-Grid infrastructure.

\paragraph{The Galaxy Portal/JSON} The P-GRADE Portal \cite{kacsuk-sipos:2005} developed at the MTA SZTAKI Computation and Automation Institute, Budapest, provides an access to grid computational and storage resources from a web browser through a specialized workflow oriented grid portal. A workflow development environment interfaces the users through this web based portal. The portal interfaces the multiple grid computing middleware and storage resources as a backend component of the environment. A highlight of P-GRADE portal is that it provides access to multiple grids and scientific communities in a single environment leading to a highly collaborative inter-community, inter-grid research environment. An older version of P-GRADE used the condor-Dagman as workflow engine while the recent version uses its own in-house developed workflow engine called Xen.
%
\paragraph{DaGue/--} Bioflow \cite{bioflow} is a query based workflow language. It is motivated by the fact that bioinformatics applications are query based applications that benefit from the complex database queries or saved database `views'. A typical Bioflow workflow pipelines a sequence of filter queries in order to get the desired results. Bioflow is mainly utilized in the area of bioinformatics and querying genome databases. Additionally, a Bioflow workflow provides mechanisms for data manipulation and updation in the existing databases. It also includes advanced features such as parallel processing of a set of mutually independent database queries. The execution model is based on the popular database query language SQL (Structured Query Language). Processes can be defined in Bioflow which are processor counterparts in workflow paradigm and similar to stored-procedures in SQL jargon.

Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical representation of bioflow workflows. A Vizbuilder workflow links to an underlying bioflow engine and provides an interface to submit the bioflow workflows for execution.
%
\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative workflow language that is inspired by the functional programming paradigm. A Martlet workflow conveniently enables distributed parameter sweep applications wherein the language intuitively defines the computation and a matching with the distributed resources can be carried out at workflow execution time. List processing applications where a sequence of application components work over an array of data in a Multi-data multi-instruction style setup is well-suited to the Martlet's model of computation.
%
\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining system for data intensive applications developed at the University of Konstanz. It is a java-based platform with a Graphical Interface that is based on eclipse (\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists of processors belonging to categories such as readers, manipulators, learners, predictors, etc. Each processor has a configuration dialog associated with it that allows user to set its parameters. Processors are connected through data and/or control connections. A highlight of knime is support for shared- and distributed-memory workflow enactment and support for parallel workflow loops and subworkflows. The loops in Knime manifest themselves in the form of subworkflows (called metanodes in Knime) by a repeated enactment of the same subworkflow.
%
\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based script oriented workflow language used to encode parallel pipelines. StarFlow handles data as well as control flow using a combination of static, dynamic and user annotations analysis. A parallelization engine uses data and control dependencies in a workflow in order to build a parallel profile of the workflow. StarFlow has been shown to be successfully interface with the SunGrid \cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}. The key feature of StarFlow is its capability to execute workflows to maximum possible parallelism. While StarFlow provides a script-based compact and semiautomatic workflow composition environment, currently it lacks a sophisticated GUI for workflow composition.

\paragraph{Galaxy/--} Galaxy \cite{blankenberg-vonkuster-etal:2010} is a bioinformatics domain SWE based framework developed at the Penn State University, Pennsilvania. It provides a webportal based environment to build complex pipelines from an available repository of tools (\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided by Galaxy using their own Galaxy server or a web-server exposing the webservice descriptions through public WSDLs. One of the major use of Galaxy is to achieve a mix and match of data analysis tools in order to analyze a vast repository of genome based database.

@@

Shown in Table \ref{tbl:scalability} is a summary of the Scalability and optimization mechanisms of the studied SWEs. The scalability features considered are through three levels of parallelism in the form of data, workflow and pipeline. The optimization features considered are the presence of mechanisms by which static and runtime optimizations are achieved.
%
% == DATA MANAGEMENT ==
\section{Data Management and Description} \label{sec:data}
Many aspects of data handling present significant diversity among the workflow
managers. Data management approaches for scientific workflows can be classified
as resource and language level. Resource level strategies include data
management at the application level of abstraction where the data is required
to be distributed, available and securely remote-accessible. It also determines
how the data will be delivered to a resource that handles the processing and
returning the processed data or results to the next processors in the workflow.
Provenance, staging, and managing heterogeneous data structures (files, arrays
etc.) are functional requirements at this level.

@@

Table \ref{tbl:datadesc} presents a summary of the data management mechanisms of the studied SWEs. Language and resource level data management mechanisms are highlighted. At the language level, data description as coupled with that of the workflow, data types and structures are considered for the SWE studied. At the resource level, the major resource types supported and their access mechanisms are highlighted. We distinguish the remote form of Grid resource types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those resource access. Other resource types such as data repositories or file catalogues are the resource types specific to a given SWE. For instance, Askalon uses a specifically configured data repository to access data and application binaries.
%Replication is one of the solution.
% == INTERFACE TO COMPUTING AND DATA INFRASTRUCTURES ==
\section{Interface to Distributed Computing Infrastructures (DCIs)} \label{sec:inter}
Workflow environments are external systems to large scale computing
infrastructures. They require an interface to connect to and use the
infrastructures. Provided the complex and distributed nature of grids, it is
not easy and practical to interact with such an infrastructure directly. Most
distributed systems have a software service layer through which it interacts
with the outside world. These interactions include accepting queries for
information and accepting requests and data for job executions. Such a layer is
typically called the \textit{middleware}. A workflow engine intending to
interact with a grid environment must provide interface to link to such
middleware layer.

@@

\section{Usability} \label{sec:usable}
A workflow environment influences the end-user comfort and convenience towards the goal of efficiently organizing and deploying a scientific experiment involving heavy usage of computational resources. However, there is no single factor that determines the usability of a SWE. A combination of several qualitative and quantitative factors play a role in determining how usable a given SWE is.
\begin{table}[ht!]
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{SWE/Language} & \textbf{Workflow Composition} & \textbf{Reusable} & \textbf{Extendable}\\
\hline
Taverna/Scufl2 & GUI & \checkmark & \checkmark\\\hline
+/BPEL & GUI & \checkmark & \checkmark \\\hline
Triana/- & GUI & \checkmark & \checkmark \\\hline
MOTEUR/Scufl & GUI & \checkmark & -- \\\hline
Kepler/MoML & GUI & \checkmark & \checkmark \\\hline
GridNexus/GXPL & GUI/Text & -- & -- \\\hline
Swift/SwiftScript & Text & -- & -- \\\hline
Askalon/AGWL & GUI & \checkmark & \checkmark \\\hline
P-GRADE/- & Web Portal & \checkmark & -- \\\hline
Vizbuilder/bioflow & GUI/Text & \checkmark & -- \\\hline
Martlet/- & Text & -- & -- \\\hline
Knime/- & GUI & \checkmark & -- \\\hline
StarFlow/- & Text & \checkmark & -- \\\hline
GWES/GWorkflowDL & GUI & \checkmark & \checkmark \\\hline
Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
Galaxy/- & Web Portal & -- & -- \\\hline
\multirow{2}{*}{MOTEUR2/} & GUI/Text & \checkmark & --\\
Gwendia~/gscript & & & \\\hline
\end{tabular}
%\caption{A summary of SWEs Usability Analysis}
\label{tbl:usability}
\end{center}
\end{table}

@@

>>>>

\begin{enumerate}
%

\item \textbf{Scalability and Optimization.} SWEs are more likely to involve
extensive usage of distributed resources than conventional programs. Owing
to the large-scale experiments, for which scientific users turn to
automation, their demands for computational power often exceeds that of a
single personal computer. Such applications are often composed of multiple,
distinct computational stages making them well-suited to the workflow
paradigm. These properties also make them ideal candidates for a
large-scale distributed computing environment. The case for suitability of
SWEs on large-scale distributed infrastructures has been well-argued upon
in~\cite{jha-katz-etal:2009}. In such a scenario, taking explicit steps to
ensure scalable and optimized performance is vital.
%

\item \textbf{Data Management.} The scale of data transfer is much higher in
SWEs from one execution site to others as well as among storage resources
while it is lower (ranging from variables to small data structures) in
programs.
%

\item \textbf{Interface to Distributed Computing Infrastructures (DCIs).}
Building SWE's interfaces to distributed systems is a complex task. The
distributed nature of resource utilization and asynchronous execution of
workflow requires a set of interdependent parameters to be configured
correctly. Further complexity is added because of complex dataflow
requirements, interactions with diverse tools and linking with legacy
application codes.

%
\item \textbf{Expressiveness.} Programming languages are inherently control
flow oriented where data flow is a secondary aspect. In fact, the
programmer must explicitly make sure that correct the inputs are available
to instructions and are in the right format. A program proceeds as the
execution of its statements from beginning to end and completes its
execution in a predefined sequence. However, in SWEs, the flow is often
determined by the data availability for execution. This model of
computation is known as dataflow computing. A workflow thus does not follow
a fixed order of execution and different execution sequences are possible
which are dominated by dataflow. The unit of work in SWEs is at a higher
level of granularity. A unit of execution is coarse-grained, and often
termed as ``processor'', ``task'', ``service'' or ``actor'' in the SWE
community. This notion of high-level granularity drives the depth to which
a workflow can be expressed in a given language form. In the rest of this
paper, this unit of work will be referred to as \textit{processor}. On the
other hand, a unit of execution in most programming languages is a much
simpler and finer executable statement that accomplishes a small
computational task. This presents a whole new set of requirements for
expressing the processors and how they should interact and operate upon
data.

%
\item \textbf{Usability.} SWEs predominantly serve as a tool for the domain
experts to utilize the computational power. This makes it essential for
them to have a friendly, and intuitive user-interface for novice and
non-expert users. For this reason, SWE user interfaces are dominated by the
visual paradigm of programming. On the other hand, traditional programs are
most often written by expert programmers. Thus, they have a text-based
composition environment albeit with visually assisting programming
environments such as IDEs. However, they are seldom to the extent that a
non-expert user can graphically compose programs.

@@

\section{Architecture of a Generic SWE}
Figure \ref{fig:wfms} shows a high-level architecture of a SWE. It shows key
components and their interaction in a SWE. These components are the focus of
interest for the present work. A \textit{user interface} interacts with the
\textit{workflow engine} using commands and messages. The \textit{workflow
language} becomes an intermediate representation for communication between user
interface and the \textit{workflow engine}. The workflow engine acts as a
compiler/interpreter and scheduler for the workflow tasks and interfaces with
the underlying \textit{distributed resources}. It also monitors the workflow
progress and collects the intermediate and final results.

%
\begin{figure}[htb]
\begin{center}
\includegraphics[width=\linewidth]{figures/workflow_interfaces_and_language}
\caption{A generic SWE scheme showing the relationship and interaction among important components}
\label{fig:wfms}
\end{center}
\end{figure}
%
SWEs are used as tools to benefit from the underlying large-scale computing
infrastructure such as clusters, supercomputers, clouds, grids. Such
environments are in use today for production level work in different scientific
domains worldwide. They do the basic task of linking inter-related processors
of a multi-staged scientific computation. However, they differ in the features
they provide based upon the scientific domain they offer service for, and the
computational model they follow. Often, one of the main reasons for such
difference is the specialized components and requirements for a particular
domain. Owing to this fact, a given workflow system dominates a particular area
of science (eg. astronomy, biology, etc.). Nonetheless, there exist a plethora
of SWEs suitable for general purpose workflow style computations.
% Ketanmark

@@

MOTEUR has sophisticated data management mechanisms within the workflow
definition. It provides a diverse set of iterations strategies well adapted for
different combinations of input datasets at the processor. An iteration
strategy allows different combinations of the data arriving at a processor
along its multiple inputs. Running legacy application code on grids has been
eased by MOTEUR by providing a generic grid-service wrapper for these
applications \cite{glatard-montagnat-etal:2008a}. Grid-services thus created
can be readily deployed on the grids by the MOTEUR workflow enactor. MOTEUR
workflows are written in the scufl language and are compatible with those
written for the Taverna SWE. It also supports an extension of scufl language
called the Gwendia language, developed at the same laboratory. 
%
\paragraph{Makeflow} A make-like workflow platform~\cite{makeflow}.

%
\paragraph{Triana/xml} Triana \cite{taylor-shields-etal:2003} is a java-based
SWE developed at the Cardiff University. It consists of a workflow composition
GUI, a framework to integrate workflow with the different execution
environments and middlewares and a set of predefined services written as java
components. A Triana workflow is expressed as an XML grammar similar to a WSDL
specification for webservices \cite{taylor-wand-etal:2005}. A highlight of
Triana is that it provides ready-to-use interface for a wide variety of
middleware and distributed infrastructures including P2P, grids and
web-services frameworks. A Triana workflow provides dataflow as well as control
flow support. Triana allows conditional branching and looping in its workflows.

%
\paragraph{Kepler/MoML} Kepler \cite{ludascher-altintas-etal:2005a} is a SWE
based upon the Ptollemy II framework \cite{eker-janneck-etal:2003} developed at
the University of California. Kepler workflows follow an actor-director
metaphor where an actor represents a unit of execution and the director
represents the paradigm for executing the workflows consisting of the actors.
Depending upon the types of directors, a Kepler workflow may exhibit a
different type of execution behavior. A workflow designer can chose from among
directors based upon the purpose of a workflow. Few of these include Process
Network (PN) for a parallel execution at the workflow level, Synchronous Data
Flow (SDF) for simple sequential data flow and Collection Oriented Modeling and
Design (COMAD) \cite{keplercomad} for a collection oriented workflow. Kepler
uses the Modeling Markup Language (MoML)~\cite{lee-neuendorffer:2000} for its
workflow representation. MoML is an extended XML syntax providing a rich
vocabulary to compose Kepler workflows. A workflow processor is called an
`actor' in Kepler.

%
\paragraph{Dryad/--} The GridNexus \cite{GridNexus} SWE, developed at the
University of North Carolina, Wilmington, facilitates a graphical composition
and execution of DAG-based workflows. Like Kepler, GridNexus is also based on
Ptolemy II framework. GridNexus workflows are expressed in an XML-based
scripting language called JXPL \cite{hunt-ferner-etal:2005}. JXPL is a
customizable and extensible scripting language allowing users to write their
own workflow processor types. GridNexus workflows are grid oriented and can be
deployed on grid resources accessed through a persistent WS-RF (Web Services
Resource Framework) interface.

%
\paragraph{Pegasus/DAX} Pegasus \cite{deelman-peg} is a low-level distributed
computing framework that can map abstract workflows as DAGs to the concrete
underlying resources using intelligent and dynamic scheduling policies. A
Pegasus workflow is expressed in a simple XML-based language called DAX (DAG in
XML). DAX describes the logical workflow components, their inputs and outputs
and their links and dependencies in a Pegasus workflow. These kind of workflows
has been shown to enact optimally using Pegasus' smart memory and disk
management techniques. Pegasus manager buffers and maintains a fix-sized window
of executable tasks before accepting new tasks.

@@

% Put this somewhere else, either in the analysis or may be in the conclusions and summary.
%The \textit{gscript} \cite{maheshwari-montagnat:2010} workflow language developed as a part of the current work is similar in spirit to the SwiftScript. Similar to SwiftScript, gscript adopts futures to express asynchronous execution in a script-based language. A SwiftScript workflow contains statically typed datasets that can be specified using ``annotations'' in the script. Unlike SwiftScript, gscript does provide user-defined strategies in order to effect data combinations in the workflow. SwiftScript provides for fault-tolerance and exception handling at run-time which is lacking currently in gscript. The underlying datasets organization in swift is represented using an XDTM XML-representation. This means that effectively, a workflow is SwiftScript plus the XDTM data representation. A workflow is simpler and compact in gscript as its data description is embedded within a workflow representation. SwiftScript supports creation of nested-workflows in the form of compound procedures. Conditional and Loop execution is supported using the if and for/while statements in both the languages.

%
\paragraph{ASKALON/AGWL.} The ASKALON Problem Solving Environment (PSE)
\cite{fahringer-prodan-etal:2007} is developed at the University of Innsbruck.
ASKALON is described as a Scientific PSE capable of composing, hosting,
executing, sharing and reusing scientific workflows written in the Abstract
Grid Workflow Language (AGWL). AGWL is an XML-based workflow language that is
capable of expressing scientific application components chains at a high-level
of abstraction. A concrete workflow consists of the abstract workflow combined
with the resources and target execution environment binding the workflow. The
graphical interface of AGWL is based upon the UML (Unified Modeling Language)
diagrams.

%
\paragraph{GWES/GWorkflowDL} The Grid Workflow Execution Service
(GWES)~\cite{Hoheisel2007} is a SWE developed by the Fraunhofer-First
institute. It implements a dynamic workflow composition and enactment model
based on the Grid Workflow Description Language (GWorkflowDL)~\cite{Alt2006}.
GWES enables key workflow development and deployment lifecycle phases including
visual composition, data-staging, Grid middleware interfacing and execution
monitoring. Currently, GWES is intensively being used in the field of Medical
Image Processing especially at the Charit\'{e} Hospital at Berlin in Germany.
It has been successfully interfaced with the German D-Grid infrastructure.
%

\paragraph{The P-GRADE Portal/--} The P-GRADE Portal \cite{kacsuk-sipos:2005}
developed at the MTA SZTAKI Computation and Automation Institute, Budapest,
provides an access to grid computational and storage resources from a web
browser through a specialized workflow oriented grid portal. A workflow
development environment interfaces the users through this web based portal. The
portal interfaces the multiple grid computing middleware and storage resources
as a backend component of the environment. A highlight of P-GRADE portal is
that it provides access to multiple grids and scientific communities in a
single environment leading to a highly collaborative inter-community,
inter-grid research environment. An older version of P-GRADE used the
condor-Dagman as workflow engine while the recent version uses its own in-house
developed workflow engine called Xen.

%
\paragraph{DaGue/--} Bioflow \cite{bioflow} is a query based workflow language.
It is motivated by the fact that bioinformatics applications are query based
applications that benefit from the complex database queries or saved database
`views'. A typical Bioflow workflow pipelines a sequence of filter queries in
order to get the desired results. Bioflow is mainly utilized in the area of
bioinformatics and querying genome databases. Additionally, a Bioflow workflow
provides mechanisms for data manipulation and updation in the existing
databases. It also includes advanced features such as parallel processing of a
set of mutually independent database queries. The execution model is based on
the popular database query language SQL (Structured Query Language). Processes
can be defined in Bioflow which are processor counterparts in workflow paradigm
and similar to stored-procedures in SQL jargon.

Vizbuilder \cite{vizbuilder} is a graphical toolkit to generate the graphical
representation of bioflow workflows. A Vizbuilder workflow links to an
underlying bioflow engine and provides an interface to submit the bioflow
workflows for execution.

%
\paragraph{*/Martlet} Martlet \cite{martlet} is a text-based, declarative
workflow language that is inspired by the functional programming paradigm. A
Martlet workflow conveniently enables distributed parameter sweep applications
wherein the language intuitively defines the computation and a matching with
the distributed resources can be carried out at workflow execution time. List
processing applications where a sequence of application components work over an
array of data in a Multi-data multi-instruction style setup is well-suited to
the Martlet's model of computation.

%
\paragraph{Knime/--} Knime \cite{knime} is a distributed, parallel pipelining
system for data intensive applications developed at the University of Konstanz.
It is a java-based platform with a Graphical Interface that is based on eclipse
(\textit{http://www.eclipse.org}) plug-in framework. A Knime workflow consists
of processors belonging to categories such as readers, manipulators, learners,
predictors, etc. Each processor has a configuration dialog associated with it
that allows user to set its parameters. Processors are connected through data
and/or control connections. A highlight of knime is support for shared- and
distributed-memory workflow enactment and support for parallel workflow loops
and subworkflows. The loops in Knime manifest themselves in the form of
subworkflows (called metanodes in Knime) by a repeated enactment of the same
subworkflow.
%
\paragraph{*/} StarFlow \cite{angelino-yamins-etal:2010} is a Python-based
script oriented workflow language used to encode parallel pipelines. StarFlow
handles data as well as control flow using a combination of static, dynamic and
user annotations analysis. A parallelization engine uses data and control
dependencies in a workflow in order to build a parallel profile of the
workflow. StarFlow has been shown to be successfully interface with the SunGrid
\cite{gentzsch:2001} Schedular and Amazon EC2 cloud infrastructure \cite{ec2}.
The key feature of StarFlow is its capability to execute workflows to maximum
possible parallelism. While StarFlow provides a script-based compact and
semiautomatic workflow composition environment, currently it lacks a
sophisticated GUI for workflow composition.

\paragraph{Galaxy Portal/JSON} Galaxy~\cite{blankenberg-vonkuster-etal:2010} is a
bioinformatics domain SWE based framework developed at the Penn State
University, Pennsilvania. It provides a webportal based environment to build
complex pipelines from an available repository of tools
(\textit{http://g2.bx.psu.edu}). Users can extend the tool repository provided
by Galaxy using their own Galaxy server or a web-server exposing the webservice
descriptions through public WSDLs. One of the major use of Galaxy is to achieve
a mix and match of data analysis tools in order to analyze a vast repository of
genome based database.

@@

Shown in Table \ref{tbl:scalability} is a summary of the Scalability and
optimization mechanisms of the studied SWEs. The scalability features
considered are through three levels of parallelism in the form of data,
workflow and pipeline. The optimization features considered are the presence of
mechanisms by which static and runtime optimizations are achieved.

%
% == DATA MANAGEMENT ==
\section{Data Management and Description} \label{sec:data}
Many aspects of data handling present significant diversity among the workflow
managers. Data management approaches for scientific workflows can be classified
as resource and language level. Resource level strategies include data
management at the application level of abstraction where the data is required
to be distributed, available and securely remote-accessible. It also determines
how the data will be delivered to a resource that handles the processing and
returning the processed data or results to the next processors in the workflow.
Provenance, staging, and managing heterogeneous data structures (files, arrays
etc.) are functional requirements at this level.

@@

Table \ref{tbl:datadesc} presents a summary of the data management mechanisms
of the studied SWEs. Language and resource level data management mechanisms are
highlighted. At the language level, data description as coupled with that of
the workflow, data types and structures are considered for the SWE studied. At
the resource level, the major resource types supported and their access
mechanisms are highlighted. We distinguish the remote form of Grid resource
types by the specific Grid protocols (\textit{eg.} GridFTP) supported for those
resource access. Other resource types such as data repositories or file
catalogues are the resource types specific to a given SWE. For instance,
Askalon uses a specifically configured data repository to access data and
application binaries.

%Replication is one of the solution.
% == INTERFACE TO COMPUTING AND DATA INFRASTRUCTURES ==
\section{Interface to Distributed Computing Infrastructures (DCIs)} \label{sec:inter}
Workflow environments are external systems to large scale computing
infrastructures. They require an interface to connect to and use the
infrastructures. Provided the complex and distributed nature of grids, it is
not easy and practical to interact with such an infrastructure directly. Most
distributed systems have a software service layer through which it interacts
with the outside world. These interactions include accepting queries for
information and accepting requests and data for job executions. Such a layer is
typically called the \textit{middleware}. A workflow engine intending to
interact with a grid environment must provide interface to link to such
middleware layer.

@@

\section{Usability} \label{sec:usable}
A workflow environment influences the end-user comfort and convenience towards
the goal of efficiently organizing and deploying a scientific experiment
involving heavy usage of computational resources. However, there is no single
factor that determines the usability of a SWE. A combination of several
qualitative and quantitative factors play a role in determining how usable a
given SWE is.

\begin{table}[ht!]
\begin{center}
% use packages: array
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{SWE/Language} & \textbf{Workflow Composition} & \textbf{Reusable} & \textbf{Extendable}\\
\hline
Taverna/Scufl2 & GUI & \checkmark & \checkmark\\\hline
+/BPEL & GUI & \checkmark & \checkmark \\\hline
Triana/- & GUI & \checkmark & \checkmark \\\hline
MOTEUR/Scufl & GUI & \checkmark & -- \\\hline
Kepler/MoML & GUI & \checkmark & \checkmark \\\hline
GridNexus/GXPL & GUI/Text & -- & -- \\\hline
Swift/SwiftScript & Text & -- & -- \\\hline
Askalon/AGWL & GUI & \checkmark & \checkmark \\\hline
P-GRADE/- & Web Portal & \checkmark & -- \\\hline
Vizbuilder/bioflow & GUI/Text & \checkmark & -- \\\hline
Martlet/- & Text & -- & -- \\\hline
Knime/- & GUI & \checkmark & -- \\\hline
StarFlow/- & Text & \checkmark & -- \\\hline
GWES/GWorkflowDL & GUI & \checkmark & \checkmark \\\hline
Pegasus/DAX & GUI & \checkmark & \checkmark \\\hline
Galaxy/- & Web Portal & -- & -- \\\hline
\multirow{2}{*}{MOTEUR2/} & GUI/Text & \checkmark & --\\
Gwendia~/gscript & & & \\\hline
\end{tabular}
%\caption{A summary of SWEs Usability Analysis}
\label{tbl:usability}
\end{center}
\end{table}

@@

% -------------------- Fit here as not main content ------------------------
